I/O simulation model with Linux page cache, integration and evaluation in SimGrid framework

### 2020-07-22 paper_io_sim.tex
Methods updated

	\section{Introduction}
		\begin{itemize}
			\item HPC, the bottleneck in I/O and the demand of HPC experiments. 
			\item Difficulties in conducting high performance computing experiments and the need of simulation frameworks.
			\item Existing experiment methods, simulators, simulation frameworks. The advantages of SimGrid compared to others \cite{casanova2008, lebre2015}. The missing of the ability to simulate page cache in SimGrid \cite{lebre2015}.
			\item The objective of the paper: Add capability to simulate I/O with page cache in SimGrid.
			\item Clarify, distinguish between memory I/O (i/o to page cache) and disk I/O.
		\end{itemize}
	\section{Related Work}

@@

		\subsection{Simulators}
			\begin{itemize}
				\item Discuss some existing methods, simulation frameworks to conduct HPC experiments. Compare pros and cons (accuracy, simulation time, usability) of some simulators (SimGrid, GridSim).
				\item Related development: RAM energy consumption \cite{gill2019} \cite{ouarnoughi2017} 
				\item What is implemented in simulators that useful for our model:
				distinguished and shared bandwidths, concurrency, device latency. 
				\item Discuss the pros of SimGrid and the reasons why we chose it to extend. (Section 2.2.2 in \cite{casanova2014})
				\item \tristan{Mention the status of storage simulation in simulators, so that we understand from Fig 1 that storage devices are available}
			\end{itemize}

@@

            We separate our simulation model in two components, the IO
			Controller and the Memory Manager, which together simulate 
			file read/write mechanisms in Linux (Figure~\ref{fig:interaction}). 
			The Memory Manager simulates I/Os to cache, flushing, and cache eviction, 
			whereas the IO Controller simulates file reads and writes. 
			Simulated applications can simply interact with the IO Manager instead of 
			calling simulated storage directly as is the case with current simulators.

@@

@@

			Next, we simulate data flushing mechanism, which writes back 
			dirty data to disk. 
			The data flushing simulation function traverses the sorted inactive first, 
			and then the active list, looks for least recently used dirty blocks, 
			writes them to disk through the storage device model, and 
			sets dirty flags of these blocks to 0 until the total flushed amount 
			reaches the amount to be flushed or there is no dirty data left in cache. 
			If the last flushed block is not entirely flushed, it is split into 
			two blocks, one that is flushed and one that remains dirty.
			The flushing time is simulated with storage model.

			For periodical flushing, our simulator searches in two page cache 
			LRU lists and flushes all the expired dirty data blocks. 
			A dirty data block is considered expired if the time from its last access 
			to current simulated time is longer than the predefined expire time. 
			Periodical flushing can be concurrent with cache read/write, 
			and sequential with disk read/write. When cache accesses and 
			periodical flushing are concurrent, flushing time is not added to simulated 
			time but the amount of flushed data is limited by flushing time. 
			When data is flushed in parallel with disk accesses, the concurrency 
			is handled by the storage model used as memory, and flushing time 
			is simulated with the amount of flushed data by this model.

			Similarly to data flushing, our cache eviction simulation function frees up 
			the page cache by traversing and deleting least recently used clean 
			data blocks in the inactive list.
			The amount of evicted data is passed in and data blocks are deleted 
			from the inactive list until total evicted data reaches the amount 
			passed in, or there is no clean block left in the list.
			If the last evicted block is not entirely evicted, the block is split, 
			and only one block is deleted.
			Our cache eviction simulation does not add up to the simulated time 
			since cache eviction time is negligible in real systems.

@@

			\begin{algorithm}\caption{File read simulation}\label{alg:read}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size (assumed to fit in memory)}
						\Desc{mm}{MemoryManager object}
						\Desc{sm}{storage simulation model}
   					\EndInput
					\State mm.flush(2*fs - mm.cached(fn) - mm.evictable\_mem) 
					\State mm.evict(2*fs - mm.cached(fn) - mm.free\_mem) 
					\State mm.periodic\_flush() \Comment{Periodically flush expired dirty data}
					\If {mm.cached(fn) $>$ 0}  \Comment{Read file cached data}
    					\State mm.cache\_read(fn)  \Comment{Update cached data}
					\EndIf
					\If {mm.cached(fn) $<$ fs} \Comment{Read uncached data from disk}
						\State sm.read(fs - mm.cached(fn)) \Comment{Simulated disk read}
						\State mm.add\_to\_cache(fs - mm.cached(fn), fn)
					\EndIf

@@

			The algorithm to simulate file reads is described in Algorithm~\ref{alg:read}.   
%			In Linux systems with page cache, data can be either read from cache 
%			or read from disk with disk bandwidth. 
%			Therefore read time can be considered as the sum of the time required to 
%			read from all devices\valerie{Can you clarify if disk+memory writes are not concurrent }, enabling us to generalize reading process as 
%			two separate phases, cache read and disk read. 
			In Linux systems with page cache, data can be either read from cache 
			or read from disk concurrently. As this concurrency is modeled in simulation 
			frameworks, we simulate a file read as follows.
			When a simulated application reads a file, it sends a file read request 
			to IO Controller. 
			First, if there is not enough memory available to store two
			copies of the file (one in cache, one in anonymous memory), the
			Memory Manager is called to flush dirty data to disk and simulate 
			flushing time with memory model (line 6).
			This flushing is complemented by eviction by the Memory Manager (line 7). 
			In lines 6-7, we assume that user applications require the amount of memory 
			equal to the file size, making total required memory twice the size of 
			the file. Apart from flushing a required amount of memory, periodical 
			flushing is then triggered to flush expired dirty data (line 8).
			Next, if the file is cached, either partially or entirely, data is read from 
			cache first (line 9). Memory Manager is called to read data from cache, 
			update the corresponding data blocks in cache (line 10).

			If file is not cached or partially cached, a disk read is required (line 12). 
			The storage model simulates the disk read with the remaining 
            amount of data (line 13), and data read from disk is added to 
            cache (line 14). The total file read time is simulated at the end of the algorithm. 
            During the entire file read, the concurrency between periodical flushing, 
            cache read and disk read is handled by storage models.

			\begin{algorithm}\caption{File write simulation}\label{alg:write}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size}
						\Desc{mm}{MemoryManager object}
						\Desc{sm}{storage simulation model}
   					\EndInput
   					\State mm.periodic\_flush()
					\State remain\_dirty = dirty\_ratio * mm.avail\_mem - mm.dirty
					\State mem\_amt = min(fs, remain\_dirty)
					\If {remain\_dirty $>$ 0} \Comment{Write with memory bandwidth}
    					\State mm.evict(mem\_amt - mm.free\_mem)
    					\State mm.write(fn, mem\_amt) 
    				\EndIf
					\If {mem\_amt $<$ fs}  \Comment{Write with disk bandwidth}
						\State mm.flush(fs - mem\_amt)  \Comment{Concurrent with disk write}
						\State mm.evict(fs - mem\_amt  - mm.free\_mem) 
						\State mm.add\_to\_cache(fn, min(fs - mem\_amt, mm.free\_mem))
						\State sm.write(fs - mem\_amt)
					\EndIf

            \valerie{I think we need to take periodic flush out as it shouldn't be added to the write time}

			Algorithm~\ref{alg:write} describes our simulation of file write. 
			Initially, periodical flushing is called to asynchronously flush expired 
			dirty data at the beginning of the write (line 6).
			During file write simulation, a file is cached in memory and also written 
			to disk. Data can be written to cache without waiting for flushing or 
			being throttled before the amount of dirty data reaches dirty\_ratio. 
			Thus, our algorithm initially checks the amount of data that can be written 
			with memory write bandwidth (line 7-8).
			If this amount is greater than 0, MemoryManager is asked to evict 
			data from cache to accommodate the written data if needed (line 10). 
			Then, cache write is simulated with memory model and data written is 
			added to page cache LRU lists (11).
			After dirty\_ratio is reached, dirty data has to be flushed before new data 
			is written to cache again. 
			We flush and evict data from cache as much as possible so that the 
			remaining data can be written to cache (line 14-15). 
			Here, MemoryManager does not guarantee that the amount of flushed 
			and evicted data is enough for the remaining data. In the case it is 
			less than the required amount and existing dirty data can not be flushed, 
			the remaining data being written to cache can be flushed and evicted right away. 
			Thus, the amount of remaining data written and kept in cache 
			after the write is limited by the amount free memory, and this amount 
			is added to cache (line 16). 
			Finally, the disk write is simulated with the remaining amount of 
			data (line 17). As in the simulation of file read, all the concurrency  
			of cache write and disk write are simulated by memory and disk models.

@@

			Having our model validated, we create simulators for different use cases 
			using original SimGrid and our SimGridExt version. 
			To simulate our pipelines and system, we leverage 
			the Workflow Management System Simulation Workbench 
			(WRENCH~\cite{wrench}), a framework  built over SimGrid.
		    We compare the results of the simulators with
			original SimGrid and extended SimGrid with the results of real
			pipelines on a real system.

>>>>

	\section{Introduction}
		\begin{itemize}
			\item HPC, the bottleneck in I/O and the demand of HPC experiments. 
			\item Difficulties in conducting high performance computing experiments and the need of simulation frameworks.
			\item Existing experiment methods, simulators, simulation frameworks. The advantages of SimGrid compared to others \cite{simgrid2008, lebre2015}. The missing of the ability to simulate page cache in SimGrid \cite{lebre2015}.
			\item The objective of the paper: Add capability to simulate I/O with page cache in SimGrid.
			\item Clarify, distinguish between memory I/O (i/o to page cache) and disk I/O.
		\end{itemize}
	\section{Related Work}

@@

		\subsection{Simulators}
			\begin{itemize}
				\item Discuss some existing methods, simulation frameworks to conduct HPC experiments. Compare pros and cons (accuracy, simulation time, usability) of some simulators (SimGrid, GridSim).
				\item Related development: RAM energy consumption \cite{gill2019} \cite{ouarnoughi2017} 
				\item What is implemented in simulators that useful for our model:
				distinguished and shared bandwidths, concurrency, device latency. 
				\item Discuss the pros of SimGrid and the reasons why we chose it to extend. (Section 2.2.2 in \cite{simgrid2014})
				\item \tristan{Mention the status of storage simulation in simulators, so that we understand from Fig 1 that storage devices are available}
			\end{itemize}

@@

            We separate our simulation model in two components, the IO
			Controller and the Memory Manager, which together simulate 
			file read/write mechanisms in Linux (Figure~\ref{fig:interaction}). 
			The Memory Manager simulates I/Os to cache, flushing, periodical flushing 
			and cache eviction, whereas the IO Controller simulates file reads and writes. 
			Simulated applications can simply interact with the IO Manager instead of 
			calling simulated storage directly as is the case with current simulators.

@@

			Memory Manager simulates two parallel threads. The first thread is the 
			main thread where activities of applications are simulated, and responsible 
			for the simulation of data flushing, cache eviction and I/Os to cache. 
			The second thread is a simulated background thread, in which Memory Manager 
			periodically searches and flushes expired dirty data in page cache LRU lists 
			to disks.

@@

			Next, we simulate data flushing, which runs on the main simulated 
			thread and writes back dirty data to disk. 
			The data flushing simulation function traverses the sorted inactive first, 
			and then the active list, looks for least recently used dirty blocks, 
			writes them to disk through the storage device model, and 
			sets dirty flags of these blocks to 0 until the total flushed amount 
			reaches the amount to be flushed or there is no dirty data left in cache. 
			If the last flushed block is not entirely flushed, it is split into 
			two blocks, one that is flushed and one that remains dirty.
			The flushing time is simulated with storage model.

			Similarly to data flushing, our cache eviction simulation runs on 
			the main thread, frees up the page cache by traversing and deleting 
			least recently used clean data blocks in the inactive list.
			The amount of evicted data is passed in and data blocks are deleted 
			from the inactive list until total evicted data reaches the amount 
			passed in, or there is no clean block left in the list.
			If the last evicted block is not entirely evicted, the block is split, 
			and only one block is deleted.
			Our cache eviction simulation does not add up to the simulated time 
			since cache eviction time is negligible in real systems.

			\begin{algorithm}\caption{Periodical flushing simulation}\label{alg:pdflush}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{in}{page cache inactive list}
        				\Desc{ac}{page cache active list}
						\Desc{t}{predefined flushing time interval}
						\Desc{sm}{storage simulation model}
   					\EndInput
   					\While{has unfinished tasks or flushing}
						\State blocks = expired dirty blocks in active and inactive list
						\State flushing\_time = sm.write(blocks)
						\State Flag blocks as clean data
						\If{flushing\_time $<$ t}
							\State sleep(t - flushing\_time)
						\EndIf
					\EndWhile
				\end{algorithmic}
			\end{algorithm}

			The periodical flushing simulation is run on a simulated background thread 
			to flush expired dirty data blocks from cache to disk.
			Similar to expired dirty pages in real Linux systems, a dirty block in our model 
			is considered expired if the time from its last access to current simulated time 
			is longer than the predefined expire time.
			The simulation algorithm of periodical flushing is an infinite loop in which 
			Memory Manager searches and flushes these expired dirty blocks to disk, 
			as described in Algorithm~\ref{alg:pdflush}. 
			The loop ends when the host does not has any remaining incomplete task 
			and all dirty data has been flushed (line 6). 
			In each repetition, Memory Manager collects expired dirty blocks in two 
			page cache LRU lists (line 7), simulates a write of this data to disk (line 8), 
			and mark these blocks as clean (line 9).
			If the time to flush data does not exceed our time interval, the thread is put 
			to sleep for the remaining time (lines 10-12). 
			Then simulation time is advanced, the algorithm finishes or continues the loop 
			based on the conditions at current simulated time.
			Because periodical flushing is simulated as a background thread, it can be 
			concurrent with disk read/write. This parallelism is handled by the 
			storage model and reflected in simulated I/O time.

@@

			\begin{algorithm}\caption{File read simulation}\label{alg:read}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size (assumed to fit in memory)}
						\Desc{mm}{MemoryManager object}
						\Desc{sm}{storage simulation model}
   					\EndInput
					\State mm.flush(2*fs - mm.cached(fn) - mm.evictable\_mem) 
					\State mm.evict(2*fs - mm.cached(fn) - mm.free\_mem) 
					\If {mm.cached(fn) $>$ 0}  \Comment{Read file cached data}
    					\State cache\_read = mm.cache\_read(fn)  \Comment{Read async}
					\EndIf
					\If {mm.cached(fn) $<$ fs} \Comment{Read uncached data from disk}
						\State disk\_read = sm.read(fs - mm.cached(fn)) \Comment{Async disk read}
						\State mm.add\_to\_cache(fs - mm.cached(fn), fn)
					\EndIf					
					\State Wait for completion of cache\_read and disk\_read 
				\end{algorithmic}
			\end{algorithm}

			The algorithm to simulate file reads is described in Algorithm~\ref{alg:read}.   
%			In Linux systems with page cache, data can be either read from cache 
%			or read from disk with disk bandwidth. 
%			Therefore read time can be considered as the sum of the time required to 
%			read from all devices\valerie{Can you clarify if disk+memory writes are not concurrent }, enabling us to generalize reading process as 
%			two separate phases, cache read and disk read. 
			In Linux systems with page cache, data can be either read from cache 
			or read from disk concurrently. As this concurrency is modeled in simulation 
			frameworks, we simulate a file read as follows.
			When a simulated application reads a file, it sends a file read request 
			to IO Controller. 
			First, if there is not enough memory available to store two
			copies of the file (one in cache, one in anonymous memory), the
			Memory Manager is called to flush dirty data to disk and simulate 
			flushing time with memory model (line 6).
			This flushing is complemented by eviction by the Memory Manager (line 7). 
			Here, we assume that user applications use the amount of memory 
			equal to the file size. 
			Next, if the file is cached, either partially or entirely, data is read from cache (line 8). 
			Memory Manager is called to simulate a asynchronous cache read, 
			update the corresponding data blocks in cache (line 9). 
			If file is not cached or partially cached, a disk read is required (line 11). 
			The storage model simulates a asynchronous disk read with the remaining 
            amount of data (line 12), and data read from disk is added to 
            cache (line 13). 
            At the end of the algorithm, we wait for the completion of two 
            asynchronous activities, the cache read in line 9 and the disk read in line 12, 
            to finish our file read and get the total simulated read time (line 15).

			\begin{algorithm}\caption{File write simulation}\label{alg:write}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size}
						\Desc{mm}{MemoryManager object}
						\Desc{sm}{storage simulation model}
   					\EndInput
					\State remain\_dirty = dirty\_ratio * mm.avail\_mem - mm.dirty
					\State mem\_amt = min(fs, remain\_dirty)
					\If {remain\_dirty $>$ 0} \Comment{Write with memory bandwidth}
    					\State mm.evict(mem\_amt - mm.free\_mem)
    					\State mm.write(fn, mem\_amt) 
    				\EndIf
					\If {mem\_amt $<$ fs}  \Comment{Write with disk bandwidth}
						\State mm.flush(fs - mem\_amt)  \Comment{Concurrent with disk write}
						\State mm.evict(fs - mem\_amt  - mm.free\_mem) 
						\State mm.add\_to\_cache(fn, min(fs - mem\_amt, mm.free\_mem))
						\State sm.write(fs - mem\_amt)
					\EndIf

@@

			Algorithm~\ref{alg:write} describes our simulation of file write. 
			During file write simulation, a file is cached in memory and also written 
			to disk. Data can be written to cache without waiting for flushing or 
			being throttled before the amount of dirty data reaches dirty\_ratio. 
			Thus, our algorithm initially checks the amount of data that can be written 
			with memory write bandwidth (line 6-7).
			If this amount is greater than 0, MemoryManager is asked to evict 
			data from cache to accommodate the written data if needed (line 10). 
			Then, cache write is simulated with memory model and data written is 
			added to page cache LRU lists (10).
			After dirty\_ratio is reached, dirty data has to be flushed before new data 
			is written to cache again. 
			We flush and evict data from cache as much as possible so that the 
			remaining data can be written to cache (line 13-14). 
			Here, MemoryManager does not guarantee that the amount of flushed 
			and evicted data is enough for the remaining data. In the case it is 
			less than the required amount and existing dirty data can not be flushed, 
			the remaining data being written to cache can be flushed and evicted right away. 
			Thus, the amount of remaining data written and kept in cache 
			after the write is limited by the amount free memory, and this amount 
			is added to cache (line 15). 
			Finally, the disk write is simulated with the remaining amount of 
			data (line 16), and the algorithm finishes with the total file write simulation time.

@@

			Having our model validated, we create simulators for different use cases 
			using original SimGrid and our SimGridExt version. 
			To simulate our pipelines and system, we leverage 
			the Workflow Management System Simulation Workbench 
			(WRENCH~\cite{wrench2018}), a framework built over SimGrid.
		    We compare the results of the simulators with
			original SimGrid and extended SimGrid with the results of real
			pipelines on a real system.
